Follow-up Samples

### 2017-09-15 submission/manuscript_R2.Rmd
Pats edits

**Background.** Despite growing evidence that members of the gut microbiota can drive tumorigenesis, little is known about what happens to it after treatment for an adenoma or carcinoma. This study tests whether treatment alters the abundance of bacterial populations associated with disease versus those with a normal colon by sequencing the 16S rRNA genes in feces of `r length(rownames(good_metaf))` individuals before and after treatment (adenoma n = `r change_summary["thetayc", "adn_n"]`), advanced adenoma n = `r change_summary["thetayc", "srn_n"]`), and carcinoma n = `r change_summary["thetayc", "crc_n"]`).

**Results.** The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"< 0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`). Although Random Forest models classified pre and post-treatment samples for each diagnosis group, there
was little overlap between the OTUs that were indicative of each treatment. Training Random Forest models on a larger cohort that contained individuals with normal colons, adenomas, advanced adenomas, and carcinomas we assessed the positive probability of having one of three diagnoses and whether the pre and post-treatment samples were classified the same. Patients who had carcinomas changed towards a microbial milieu that resembles the normal colon after treatment (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05). There was no significant difference in the microbiota of individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).

**Conclusions.** It was difficult to identify significant differences between pre and post-treatment samples from patients with adenoma and advanced adenoma. However, for carcinomas, not only were there large changes in pre versus post-treatment samples, but these changes were towards a more normal microbiota.

@@

***The treatment of lesions are not consistent across diagnosis groups.***
We used two approaches to identify those bacterial populations that change between the two samples for each diagnosis group. First, we sought to identify individual OTUs that could account for the change in overall community structure. However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) > 0.05, invisible(">"), invisible("<"))` 0.05). It is likely that high inter-individual variation and the irregular distribution of OTUs across individuals limited the statistical power of the test. We attempted to overcome these problems by using Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups. The models overall performed moderately, with the carcinoma model being the best (adenoma AUC range = `r format(1 - max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(1 - min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`). Interestingly, the top 10 most important OTUs by MDA that were used for each model had little overlap with each other [Figure 2]. Although treatment had an impact on the overall community structure, the effect of treatment was not consistent across patients and diagnosis groups. Both the adenoma and carcioma treatment models had AUCs that were significantly higher than a random model permutation (P-value `r ifelse(filter(treat_random_pvalues, study == "adn")[, "pvalue"] > 0.0001 & filter(treat_random_pvalues, study == "crc")[, "pvalue"] > 0.0001, invisible(">"), invisible("<"))` 0.0001).

@@

***Model Building.*** The Random Forest [@breiman_random_2001] algorithm was used to create the three models used to classify pre and post-treatment samples by diagnosis (adenoma, advanced adenoma, or carcinoma). The total number of individuals in the pre versus post-treatment models was `r length(rownames(good_metaf))` individuals.  There were a total of `r length(rownames(filter(good_metaf, Dx_Bin == "adenoma")))` individuals in the pre versus post-treatment adenoma model, `r length(rownames(filter(good_metaf, Dx_Bin == "adv_adenoma")))` individuals in the pre versus post-treatment advanced adenoma model, and `r length(rownames(filter(good_metaf, Dx_Bin == "cancer")))` individuals in the pre versus post-treatment carcinoma model [Table 1].

Similarly, the Random Forest [@breiman_random_2001] algorithm was also used to create the three models used to classify normal versus diagnosis. These samples were obtained using the same methodology as described earlier in this section. All samples used for this component of model training were from pre-treatment samples only. The total number of individuals in the normal versus diagnosis models was `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals [Table 2]. There were a total of `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Adenoma" | dx == "normal"))[, "sample"])` individuals in the normal versus adenoma model, `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "adv Adenoma" | dx == "normal"))[, "sample"])` individuals in the normal versus advanced adenoma model, and `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Cancer" | dx == "normal"))[, "sample"])` individuals in the normal versus carcinoma model [Table 2].

All models included only OTU data obtained from 16S rRNA sequencing and were processed and cleaned using the R package caret (v`r paste(packageVersion("caret"))`). Optimization of the mtry hyper-parameter involved making 100 different 80/20 (train/test) splits of the data where the same proportion was present within both the whole data set and the 80/20 split. For each of the different splits, 20 repeated 10-fold cross validation was performed on the 80% component to optimize the mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the Receiver Operator Characteristic). The resulting model was then tested on the hold out data obtained from the 20% component. For adenoma pre versus post-treatment models the optimized mtrys were `r names(table(adn_treat_AUC_summary$best_mtry))[1]` (n = `r table(adn_treat_AUC_summary$best_mtry)[1]`), `r names(table(adn_treat_AUC_summary$best_mtry))[2]` (n = `r table(adn_treat_AUC_summary$best_mtry)[2]`), and `r names(table(adn_treat_AUC_summary$best_mtry))[3]` (n = `r table(adn_treat_AUC_summary$best_mtry)[3]`) for adenoma. For advanced adenoma the mtrys were  `r names(table(srn_treat_AUC_summary$best_mtry))[1]` (n = `r table(srn_treat_AUC_summary$best_mtry)[1]`), `r names(table(srn_treat_AUC_summary$best_mtry))[2]` (n = `r table(srn_treat_AUC_summary$best_mtry)[2]`), and `r names(table(srn_treat_AUC_summary$best_mtry))[3]` (n = `r table(srn_treat_AUC_summary$best_mtry)[3]`). For carcinoma the mtrys were  `r names(table(crc_treat_AUC_summary$best_mtry))[1]` (n = `r table(crc_treat_AUC_summary$best_mtry)[1]`), `r names(table(crc_treat_AUC_summary$best_mtry))[2]` (n = `r table(crc_treat_AUC_summary$best_mtry)[2]`), and `r names(table(crc_treat_AUC_summary$best_mtry))[3]` (n = `r table(crc_treat_AUC_summary$best_mtry)[3]`). For the normal versus diagnosis models the optimized mtrys were `r names(table(adn_AUC_run_summary$best_mtry))[1]` (n = `r table(adn_AUC_run_summary$best_mtry)[1]`), `r names(table(adn_AUC_run_summary$best_mtry))[2]` (n = `r table(adn_AUC_run_summary$best_mtry)[2]`), and `r names(table(adn_AUC_run_summary$best_mtry))[3]` (n = `r table(adn_AUC_run_summary$best_mtry)[3]`) for adenoma. For advanced adenoma `r names(table(srn_AUC_run_summary$best_mtry))[1]` (n = `r table(srn_AUC_run_summary$best_mtry)[1]`), `r names(table(srn_AUC_run_summary$best_mtry))[2]` (n = `r table(srn_AUC_run_summary$best_mtry)[2]`), and `r names(table(srn_AUC_run_summary$best_mtry))[3]` (n = `r table(srn_AUC_run_summary$best_mtry)[3]`). For carcinoma `r names(table(crc_AUC_run_summary$best_mtry))[1]` (n = `r table(crc_AUC_run_summary$best_mtry)[1]`), `r names(table(crc_AUC_run_summary$best_mtry))[2]` (n = `r table(crc_AUC_run_summary$best_mtry)[2]`), and `r names(table(crc_AUC_run_summary$best_mtry))[3]` (n = `r table(crc_AUC_run_summary$best_mtry)[3]`). The hyper-parameter, mtry, defines the number of variables to investigate at each split before a new division of the data was created with the Random Forest model [@breiman_random_2001].

The treatment models were then assessed for their strength to classify whether individuals had obtained treatment. We also assessed whether the models performed better than random by creating 100 permutations of a data set in which the labels were randomly assigned. To assess similarity of the treatment models we looked for overlap between their top 10 most important OTUs.

For the diagnosis models that were trained on a completely independent sample set, each model was then applied to our `r length(rownames(good_metaf))`-person cohort [Table 1] based on diagnosis: adenoma (pre-treatment adenoma (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "EDRN"])`) versus post-treatment adenoma (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "followUp"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "followUp"])`)), advanced adenoma pre-treatment advanced adenoma (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "EDRN"])` ) versus post-treatment advanced adenoma (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "followUp"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "followUp"])`), and carcinoma (pre-treatment carcinoma (carcinoma n = `r length(filter(good_metaf, Dx_Bin == "cancer")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "EDRN"])`) versus post-treatment carcinoma (carcinoma n = `r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free == "n")))` and disease free n = `r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free != "n")))`)). The application of the pre versus post-treatment models generated the probabilities that the sample was a pre-treatment sample.  The application of the normal versus diagnosis models generated the probabilities that the sample was that specific diagnosis (adenoma, advanced adenoma, or carcinoma). We also looked at similarities between the models by looking for overlap in the top 10% of important OTUs to each respective model.

>>>>

Despite growing evidence that members of the gut microbiota can drive tumorigenesis, little is known about how treatment impacts the colonic community. This study tested the hypothesis that treatment for adenoma or carcinoma alters the abundance of bacterial populations associated with disease to those associated with a normal colon. We tested this hypothesis by sequencing the 16S rRNA genes in the feces of `r length(rownames(good_metaf))` individuals before and after treatment for adenoma (N = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (N = `r change_summary["thetayc", "srn_n"]`), and carcinoma (N = `r change_summary["thetayc", "crc_n"]`). The structure of the microbiota in all three groups changed following treatment (P`r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"<0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`). To determine whether individuals who underwent treatment were more likely to have OTUs associated with normal colons we used a larger cohort that contained individuals with normal colons (N=XXX) and untreated individuals with adenomas (N=XXX), advanced adenomas (N=XXX), and carcinomas (N=XXX). Using these data, we built Random Forest models and measured the change in the positive probability of having one of the three diagnoses to assess whether the post-treatment samples received the same classification as the pre-treatment samples. The microbiota of patients diagnosed with carcinomas changed towards that of a normal colon following treatment (P<0.XXX). Those were initially diagnosed as having adenomas or advanced adenomas were no more likely to have a microbiota resembling that of a normal colon. These results further illuminate our understanding of the role of the gut microbiota in tumorigenesis and demonstrate the potential for using the gut microbiota as biomarkers for assessing risk and the effectiveness of treatment.

## Importance

Colorectal cancer is a worldwide health problem. Previous work in animal models and human populations has demonstrated that the gut microbiota has an important role in promoting tumorigenesis through a variety of mechanisms. Less well understood is how the gut microbiota changes following treatment for varying stages of tumorigenesis. We observed that following treatment of patients diagnosed with carcinoma, their gut microbiota is likely revert to a state associated with individuals who had normal colons whereas those diagnosed with adenomas and advanced adenomas were not. This study was a necessary step to ultimately investigating how the microbiota changes in those with carcinoma recurrence and predicting which patients are more likely to recur. Finding better strategies to stratify patients based on recurrence risk could provide a more effective means of planning therapies and monitoring patients after treatment.

@@

***The treatment of lesions are not consistent across diagnosis groups.***
We used two approaches to identify those bacterial populations that change between the two samples for each diagnosis group. First, we sought to identify individual OTUs that could account for the change in overall community structure. However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) > 0.05, invisible(">"), invisible("<"))` 0.05). It is likely that high inter-individual variation and the irregular distribution of OTUs across individuals limited the statistical power of the test. We attempted to overcome these problems by using Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups. The adenoma and carcinoma models performed well (adenoma AUC range = `r format(1 - max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(1 - min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` and carcinoma AUC range = `r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`); however, the model for patients treated for advanced adenomas was not able to reliably differentiate between the pre and post-treatment samples (advanced adenoma AUC range = `r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`). Interestingly, the top 10 most important OTUs by MDA that were used for each model had little overlap with each other [Figure 2]. Although treatment had an impact on the overall community structure, the effect of treatment was not consistent across patients and diagnosis groups. Both the adenoma and carcioma treatment models had AUCs that were significantly higher than a random model permutation (P-value `r ifelse(filter(treat_random_pvalues, study == "adn")[, "pvalue"] > 0.0001 & filter(treat_random_pvalues, study == "crc")[, "pvalue"] > 0.0001, invisible(">"), invisible("<"))` 0.0001).

@@

***Model Building.*** The Random Forest [@breiman_random_2001] algorithm was used to create the three models used to classify pre and post-treatment samples by diagnosis (adenoma, advanced adenoma, or carcinoma) as well as to assess the probability that a sample was more similar to the patient's original diagnosis or that of a disease-free patient. All models included only OTU data obtained from 16S rRNA sequencing and were processed using the caret (v`r paste(packageVersion("caret"))`) R package. For each model we optimized the mtry hyper-parameter, which defines the number of OTUs to investigate at each split before a new division of the data was created with the Random Forest model [@breiman_random_2001]. To insure that our optimization did not result in over-fitting of the data, we made 100 different 80/20 (train/test) splits of the data where the same proportion was present within both the whole data set and the 80/20 split. For each of the 100 splits, 20 repeated 10-fold cross validation was performed on the 80% component to optimize the mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the Receiver Operator Characteristic). The resulting model was then tested on the 20% of the data that were held out. A summary of the mtry hyperparameter values that were tried is available in Table SX. The reported P-values for each model relative to a random labeling was assessed by comparing the distribution of the 100 80/20 splits for the correctly labeled data to the distribution of randomly labeled data.

The three diagnosis models were constructed by using the data from Baxter et al. [@baxter_microbiota-based_2016], which was censored for the pre-treatment samples of the patients that we had post-treatment samples. The treatment models were then used to quantify the model probability that a patient with an initial diagnosis retained that diagnosis or a disease-free diagnosis.
